---
title: "Using Rule Learners to Identify Poisonous Mushrooms"
author: "Ignacio Faria"
date: "April 30, 2016"
output: pdf_document
---

We will examine a dataset of 8,124 mushrooms and their features to determine if we can predict if a mushroom is either poisonous or edible.

### Step 1 - Collecting Data ---

```{r Data input, echo=TRUE}
setwd("~/")
mushrooms <- read.csv("mushrooms.csv", stringsAsFactors = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```

### Step 2 - Exploring and Preparing the Data ---

Now that we have the data, we must observe how it is set up. Our most important variable in this dataset is mushroom type; is the mushroom poisonous or edible? Rule learners work best when the levels are evenly distributed.

```{r Exploring the data, echo=FALSE}
str(mushrooms)
mushrooms$veil_type <- NULL
table(mushrooms$type)
```

We observed that "veil_type" had one level, so it was worth omitting and also that the mushrooms edibility were also close to even.

### Step 3 - Training a Model on the Data ---

To develop this model, we will use a 1R (One Rule) algorithm to train this dataset.

```{r Training the dataset, echo=TRUE}
library(RWeka)
mushroom_1R <- OneR(type ~ ., data = mushrooms)
```

### Step 4 - Evaluating Model Performance ---

```{r Evaluating the model, echo=TRUE}
summary(mushroom_1R)
```
The one rule learner worked well.  It was accurate 98.52% of the time and only labeled 120 edible mushrooms as poisonous.  

### Step 5 - Improving Model Performance ---

This time we will apply a RIPPER rule learning algorithm and examine its performance.

```{r Applying RIPPER, echo=TRUE}
mushroom_JRip <- JRip(type ~ ., data = mushrooms)
mushroom_JRip
summary(mushroom_JRip)
```

The algorithm used 8 rules and worked with perfect accuracy.
